---
title: "NicheNet Circos Diagnostic"
output: html_document
---

# Setup

Config values are pulled directly from `config.yml` and `pipeline-config.yml`.
Run chunks sequentially to diagnose issues.

```{r setup}
library(yaml)
library(Seurat)
library(dplyr)
library(ggplot2)
library(tibble)
library(nichenetr)
library(presto)
library(RColorBrewer)
library(glue)
library(circlize)

configs         <- yaml::read_yaml("../config.yml")
pipeline_configs <- yaml::read_yaml("../pipeline-config.yml")

source("helper-functions/helper-functions.R")
```

# Config values in use

```{r config-values}
# --- Paths ---
rds_path      <- configs$paths$RDS_File
default_assay <- configs$paths$DefaultAssay
cell_type_col <- configs$paths$CellTypeCol

# --- NicheNet params ---
nn_cfg       <- configs$nichenet
org          <- nn_cfg$organism
lfc_cutoff   <- nn_cfg$lfc_cutoff
expr_pct     <- nn_cfg$expression_pct
top_n_lig    <- nn_cfg$top_n_ligands
top_n_rec    <- nn_cfg$top_n_receptors_per_ligand
top_n_tgt    <- nn_cfg$top_n_targets

# --- Niches ---
niches <- lapply(nn_cfg$niches, function(n) list(sender = n$sender, receiver = n$receiver))

# --- Database paths ---
lr_network_path         <- pipeline_configs$nichenet$paths[[org]]$lr_network
ligand_target_matrix_path <- pipeline_configs$nichenet$paths[[org]]$ligand_target_matrix

cat("RDS:            ", rds_path, "\n")
cat("Assay:          ", default_assay, "\n")
cat("Cell type col:  ", cell_type_col, "\n")
cat("Organism:       ", org, "\n")
cat("lfc_cutoff:     ", lfc_cutoff, "\n")
cat("expression_pct: ", expr_pct, "\n")
cat("top_n_ligands:  ", top_n_lig, "\n")
cat("top_n_receptors:", top_n_rec, "\n")
cat("top_n_targets:  ", top_n_tgt, "\n")
cat("\nNiches:\n")
for (nm in names(niches)) {
  cat(" ", nm, "— sender:", paste(niches[[nm]]$sender, collapse=", "),
      "| receiver:", niches[[nm]]$receiver, "\n")
}
```

# 1. Load Seurat object

```{r load-data}
stopifnot("RDS file not found" = file.exists(rds_path))
seu_obj <- readRDS(rds_path)
DefaultAssay(seu_obj) <- default_assay
seu_obj <- JoinLayers(seu_obj, assay = default_assay)
Idents(seu_obj) <- seu_obj@meta.data[[cell_type_col]]

cat("Cell types present:\n")
print(sort(unique(Idents(seu_obj))))
```

```{r check-niche-cell-types}
# Verify all sender/receiver names exist in the Seurat object
all_celltypes <- as.character(unique(Idents(seu_obj)))
for (nm in names(niches)) {
  missing_senders   <- setdiff(niches[[nm]]$sender,   all_celltypes)
  missing_receivers <- setdiff(niches[[nm]]$receiver, all_celltypes)
  if (length(missing_senders) > 0)
    warning(glue("Niche '{nm}': senders not found in data: {paste(missing_senders, collapse=', ')}"))
  if (length(missing_receivers) > 0)
    warning(glue("Niche '{nm}': receiver not found in data: {paste(missing_receivers, collapse=', ')}"))
}
cat("Cell type check complete.\n")
```

# 2. Load NicheNet databases

```{r load-databases}
stopifnot("lr_network file not found"           = file.exists(lr_network_path))
stopifnot("ligand_target_matrix file not found" = file.exists(ligand_target_matrix_path))

lr_network           <- readRDS(lr_network_path)
ligand_target_matrix <- readRDS(ligand_target_matrix_path)

lr_network <- lr_network %>%
  dplyr::rename(ligand = from, receptor = to) %>%
  distinct(ligand, receptor)

cat("lr_network rows:           ", nrow(lr_network), "\n")
cat("ligand_target_matrix dims: ", dim(ligand_target_matrix), "\n")
```

# 3. Differential expression

```{r de-senders-receivers}
DE_sender <- calculate_niche_de(
  seurat_obj = seu_obj, niches = niches, type = "sender", assay_oi = default_assay)
DE_sender_processed <- process_niche_de(
  DE_table = DE_sender, niches = niches, expression_pct = expr_pct, type = "sender")

DE_receiver <- calculate_niche_de(
  seurat_obj = seu_obj, niches = niches, type = "receiver", assay_oi = default_assay)
DE_receiver_processed <- process_niche_de(
  DE_table = DE_receiver, niches = niches, expression_pct = expr_pct, type = "receiver")

DE_sender_receiver <- combine_sender_receiver_de(
  DE_sender_processed, DE_receiver_processed, lr_network, specificity_score = "min_lfc")

cat("DE_sender rows:          ", nrow(DE_sender), "\n")
cat("DE_receiver rows:        ", nrow(DE_receiver), "\n")
cat("DE_sender_receiver rows: ", nrow(DE_sender_receiver), "\n")
```

# 4. Receiver target DE + geneset sizes

```{r geneset-sizes}
DE_receiver_targets <- calculate_niche_de_targets(
  seurat_obj    = seu_obj,
  niches        = niches,
  expression_pct = expr_pct,
  lfc_cutoff    = lfc_cutoff,
  assay_oi      = default_assay
)
DE_receiver_processed_targets <- process_receiver_target_de(
  DE_receiver_targets = DE_receiver_targets,
  niches              = niches,
  expression_pct      = expr_pct,
  specificity_score   = "min_lfc"
)

# Check geneset sizes — vignette recommends 20–1000 genes; adjust lfc_cutoff if outside range
for (niche_oi in names(niches)) {
  receiver_oi <- niches[[niche_oi]]$receiver
  geneset <- DE_receiver_processed_targets %>%
    filter(receiver == receiver_oi & niche == niche_oi & target_score > lfc_cutoff) %>%
    pull(target) %>% unique()
  cat(glue("Niche '{niche_oi}' geneset size: {length(geneset)}"), "\n")
  if (length(geneset) < 20)
    warning(glue("Geneset for '{niche_oi}' has < 20 genes — consider lowering lfc_cutoff in config.yml"))
  if (length(geneset) > 1000)
    warning(glue("Geneset for '{niche_oi}' has > 1000 genes — consider raising lfc_cutoff in config.yml"))
}
```

# 5. Ligand activity

```{r ligand-activity}
niche_geneset_list <- lapply(names(niches), function(niche_oi) {
  receiver_oi <- niches[[niche_oi]]$receiver
  geneset <- DE_receiver_processed_targets %>%
    filter(receiver == receiver_oi & niche == niche_oi & target_score > lfc_cutoff) %>%
    pull(target) %>% unique()
  background <- rownames(seu_obj)[
    Matrix::rowSums(GetAssayData(seu_obj, assay = default_assay)) > 0]
  list(receiver = receiver_oi, geneset = geneset, background = background)
})
names(niche_geneset_list) <- names(niches)

ligand_activities_targets <- get_ligand_activities_targets(
  niche_geneset_list   = niche_geneset_list,
  ligand_target_matrix = ligand_target_matrix,
  top_n_target         = top_n_tgt
)

cat("ligand_activities_targets rows:", nrow(ligand_activities_targets), "\n")
```

# 6. Expression tables + prioritization

```{r expression-and-prioritization}
prioritizing_weights <- c(
  scaled_ligand_score                        = 5,
  scaled_ligand_expression_scaled            = 1,
  ligand_fraction                            = 1,
  scaled_ligand_score_spatial                = 0,
  scaled_receptor_score                      = 0.5,
  scaled_receptor_expression_scaled          = 0.5,
  receptor_fraction                          = 1,
  ligand_scaled_receptor_expression_fraction = 1,
  scaled_receptor_score_spatial              = 0,
  scaled_activity                            = 0,
  scaled_activity_normalized                 = 1
)

features_oi <- union(lr_network$ligand, lr_network$receptor) %>%
  union(ligand_activities_targets$target) %>% setdiff(NA)

dotplot <- suppressWarnings(Seurat::DotPlot(
  seu_obj %>% subset(idents = niches %>% unlist() %>% unique()),
  features = features_oi, assay = default_assay
))
exprs_tbl <- dotplot$data %>% as_tibble() %>%
  rename(celltype = id, gene = features.plot, expression = avg.exp,
         expression_scaled = avg.exp.scaled, fraction = pct.exp) %>%
  mutate(fraction = fraction / 100) %>%
  select(celltype, gene, expression, expression_scaled, fraction) %>%
  distinct() %>% arrange(gene) %>% mutate(gene = as.character(gene))

exprs_tbl_ligand <- exprs_tbl %>%
  filter(gene %in% lr_network$ligand) %>%
  rename(sender = celltype, ligand = gene, ligand_expression = expression,
         ligand_expression_scaled = expression_scaled, ligand_fraction = fraction) %>%
  mutate(scaled_ligand_expression_scaled = scale_quantile_adapted(ligand_expression_scaled)) %>%
  mutate(ligand_fraction_adapted = ligand_fraction) %>%
  mutate_cond(ligand_fraction >= expr_pct, ligand_fraction_adapted = expr_pct) %>%
  mutate(scaled_ligand_fraction_adapted = scale_quantile_adapted(ligand_fraction_adapted))

exprs_tbl_receptor <- exprs_tbl %>%
  filter(gene %in% lr_network$receptor) %>%
  rename(receiver = celltype, receptor = gene, receptor_expression = expression,
         receptor_expression_scaled = expression_scaled, receptor_fraction = fraction) %>%
  mutate(scaled_receptor_expression_scaled = scale_quantile_adapted(receptor_expression_scaled)) %>%
  mutate(receptor_fraction_adapted = receptor_fraction) %>%
  mutate_cond(receptor_fraction >= expr_pct, receptor_fraction_adapted = expr_pct) %>%
  mutate(scaled_receptor_fraction_adapted = scale_quantile_adapted(receptor_fraction_adapted))

exprs_tbl_target <- exprs_tbl %>%
  filter(gene %in% ligand_activities_targets$target) %>%
  rename(receiver = celltype, target = gene, target_expression = expression,
         target_expression_scaled = expression_scaled, target_fraction = fraction)

exprs_sender_receiver <- lr_network %>%
  inner_join(exprs_tbl_ligand, by = c("ligand")) %>%
  inner_join(exprs_tbl_receptor, by = c("receptor")) %>%
  inner_join(DE_sender_receiver %>% distinct(niche, sender, receiver))

ligand_scaled_receptor_expression_fraction_df <- exprs_sender_receiver %>%
  group_by(ligand, receiver) %>%
  mutate(rank_receptor_expression = dense_rank(receptor_expression),
         rank_receptor_fraction   = dense_rank(receptor_fraction)) %>%
  mutate(ligand_scaled_receptor_expression_fraction = 0.5 * (
    (rank_receptor_fraction  / max(rank_receptor_fraction)) +
    (rank_receptor_expression / max(rank_receptor_expression))
  )) %>%
  distinct(ligand, receptor, receiver, ligand_scaled_receptor_expression_fraction) %>%
  ungroup()

spatial_info <- tibble(celltype_region_oi = NA, celltype_other_region = NA,
                       niche = NA, celltype_type = NA)
sender_spatial_DE_processed <- get_non_spatial_de(
  niches = niches, spatial_info = spatial_info, type = "sender", lr_network = lr_network) %>%
  mutate(scaled_ligand_score_spatial = scale_quantile_adapted(ligand_score_spatial))

receiver_spatial_DE_processed <- get_non_spatial_de(
  niches = niches, spatial_info = spatial_info, type = "receiver", lr_network = lr_network) %>%
  mutate(scaled_receptor_score_spatial = scale_quantile_adapted(receptor_score_spatial))

output <- list(
  DE_sender_receiver                          = DE_sender_receiver,
  ligand_scaled_receptor_expression_fraction_df = ligand_scaled_receptor_expression_fraction_df,
  sender_spatial_DE_processed                 = sender_spatial_DE_processed,
  receiver_spatial_DE_processed               = receiver_spatial_DE_processed,
  ligand_activities_targets                   = ligand_activities_targets,
  DE_receiver_processed_targets               = DE_receiver_processed_targets,
  exprs_tbl_ligand                            = exprs_tbl_ligand,
  exprs_tbl_receptor                          = exprs_tbl_receptor,
  exprs_tbl_target                            = exprs_tbl_target
)

prioritization_tables <- get_prioritization_tables(output, prioritizing_weights)
cat("prioritization_tbl_ligand_receptor rows:", nrow(prioritization_tables$prioritization_tbl_ligand_receptor), "\n")
cat("prioritization_tbl_ligand_target rows:  ", nrow(prioritization_tables$prioritization_tbl_ligand_target), "\n")
```

# 7. Circos plot diagnostic

This section isolates exactly what the pipeline does so you can see whether the
PDF is written correctly. The fix is that `pdf()` must be opened **before**
calling `make_circos_lr` — circlize draws to the active device as a side effect.

```{r circos-test}
top_ligand_receptor_niche_df <- prioritization_tables$prioritization_tbl_ligand_receptor %>%
  select(niche, sender, receiver, ligand, receptor, prioritization_score) %>%
  group_by(ligand, receptor) %>% top_n(1, prioritization_score) %>%
  ungroup() %>% select(ligand, receptor, niche) %>%
  rename(top_niche = niche)

for (niche_oi in names(niches)) {
  receiver_oi <- niches[[niche_oi]]$receiver

  filtered_ligands <- prioritization_tables$prioritization_tbl_ligand_receptor %>%
    filter(receiver == receiver_oi) %>%
    top_n(top_n_lig, prioritization_score) %>%
    pull(ligand) %>% unique()

  cat(glue("\nNiche '{niche_oi}': {length(filtered_ligands)} top ligands"), "\n")

  prioritized_tbl_oi <- prioritization_tables$prioritization_tbl_ligand_receptor %>%
    filter(ligand %in% filtered_ligands) %>%
    select(niche, sender, receiver, ligand, receptor, ligand_receptor, prioritization_score) %>%
    distinct() %>%
    inner_join(top_ligand_receptor_niche_df) %>%
    group_by(ligand) %>%
    filter(receiver == receiver_oi) %>%
    top_n(top_n_rec, prioritization_score) %>%
    ungroup()

  cat(glue("  prioritized_tbl_oi rows: {nrow(prioritized_tbl_oi)}"), "\n")

  if (nrow(prioritized_tbl_oi) == 0) {
    warning(glue("No LR pairs to plot for niche '{niche_oi}' — skipping circos"))
    next
  }

  n_senders <- length(unique(prioritized_tbl_oi$sender))
  cat(glue("  unique senders: {n_senders}"), "\n")

  # brewer.pal requires n >= 3; if fewer senders, pad with extra colors
  pal_n <- max(3, n_senders)
  colors_sender <- brewer.pal(n = pal_n, name = "Spectral")[seq_len(n_senders)] %>%
    magrittr::set_names(prioritized_tbl_oi$sender %>% unique() %>% sort())

  colors_receiver <- c("lavender") %>%
    magrittr::set_names(prioritized_tbl_oi$receiver %>% unique() %>% sort())

  out_path <- glue("../lrp-outs/niche-net/plots/{niche_oi}-circos-diag.pdf")
  dir.create(dirname(out_path), recursive = TRUE, showWarnings = FALSE)

  # pdf() MUST be opened before make_circos_lr — circlize draws to the active device
  pdf(out_path, width = 10, height = 10)
  tryCatch({
    make_circos_lr(prioritized_tbl_oi, colors_sender, colors_receiver)
    cat(glue("  PDF written: {out_path}"), "\n")
  }, error = function(e) {
    cat(glue("  ERROR in make_circos_lr: {e$message}"), "\n")
  })
  dev.off()
}
```

# Session info

```{r session-info}
sessionInfo()
```
